---
title: "HW 4 Missing Data & Model Selection: Data Analysis Problems "
subtitle: "Advanced Regression (STAT 353-0)"
author: "YOUR NAME"
pagetitle: "HW 4 YOUR NAME"
date: today

format:
  html:
    toc: true
    toc-depth: 4
    toc-location: left
    embed-resources: true
    code-fold: false
    link-external-newwindow: true
    theme: cosmo

execute:
  warning: false

from: markdown+emoji
reference-location: margin
citation-location: margin
---

::: {.callout-tip icon=false}

## Github Repo Link

A link to your GitHub repo should be placed at the top your submitted html.

:::

::: {.callout-important}

All students are required to complete this problem set!

:::


## Data analysis problems

### 1. Exercise D20.1 (MI)

Using the United Nations social-indicators data (in `UnitedNations.txt`), develop a regression model for the response variable female expectation of life. Feel free to use whatever explanatory variables in the data set make sense to you and to employ variable transformations, if needed.

(a) Work initially with complete cases, and once you have an apparently satisfactory model, obtain estimates and standard errors of the regression coefficients.

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::

(b) Now redo your analysis in (a) but use multiple imputation.

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::

(c) Compare these results to those from the complete-case analysis. What do you conclude?

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::


### 2. Exercise D20.3 (Selection)

Long (1997) reports a regression in which the response variable is the prestige of the academic departments where PhDs in biochemistry find their first jobs. The data are in the file `Long-PhDs.txt`.

Prestige is measured on a scale that runs from 1.00 to 5.00, and is unavailable for departments without graduate programs and for departments with ratings below 1.00. The explanatory variables include a dummy regressor for gender; the prestige of the department in which the individual obtained his or her PhD; the number of citations received by the individualís mentor; a dummy regressor coding whether or not the individual held a fellowship; the number of articles published by the individual; and the number of citations received by the individual.

Estimate the regression of prestige of first job on the other variables in three ways:

(a) code all of the missing values as 1.00 and perform an OLS regression;

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::

(b) treat the missing values as truncated at 1.00 and employ Heckmanís selection-regression model;

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::

(c) treat the missing values as censored and fit the Tobit model.

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::

(d) Compare the estimates and coefficient standard errors obtained by the three approaches. Which of these approaches makes the most substantive sense?

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::

### 3. Exercise (Bootstrap)

We will now consider the `Boston` housing dataset from the `MASS` package.

```{r}
#| label: load-boston-data

# load Boston data
data(Boston, package = "MASS")
```

Run `??Boston` in condole to see codebook.

(a) Provide an estimate of the population mean of `medv`. Call this estimate $\hat{\mu}$.

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::

(b) What is the formula for the standard error of an estimate of the mean? Use this to provide an estimate of the standard error of $\hat{\mu}$ in (a).

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::

(c) Estimate this standard error using the bootstrap. How does this compare to the answer from (b)?

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::

(d) Provide an estimate of $\hat{\mu}_{med}$, the  median value of `medv` in the population.

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::

(e) Estimate the standard error of $\hat{\mu}_{med}$. Notice that there is no simple formula to do this, so instead use the bootstrap. Comment on your findings.

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::

### 4. Exercise D22.1 (Model Selection)

The data file `BaseballPitchers.txt` contains salary and performance data for major-league baseball pitchers at the start of the 1987 season. The data are analogous to those for baseball hitters used as an example in the chapter. Be sure to explore the data and think about variables to use as predictors before specifying candidate models.

(a) Employing one or more of the methods of model selection described in the text, develop a regression model to predict pitchers' salaries.

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::

(b) How successful is the model in predicting salaries? Does the model make substantive sense?

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::