---
title: "HW 2: Generalized Linear Models"
subtitle: "Advanced Regression (STAT 353-0)"
author: "YOUR NAME"
pagetitle: "HW 2 YOUR NAME"
date: today

format:
  html:
    toc: true
    toc-depth: 4
    toc-location: left
    embed-resources: true
    code-fold: false
    link-external-newwindow: true
    theme: cosmo

execute:
  warning: false

from: markdown+emoji
reference-location: margin
citation-location: margin  
---

::: {.callout-tip icon=false}

## Github Repo Link

To link to your github **repo**sitory, appropriately edit the example link below. Meaning replace `https://your-github-repo-url` with your github repo url. Suggest verifying the link works before submitting.

[https://your-github-repo-url](https://your-github-repo-url)

:::

## Handwork

Policy on handwork is that it is required for Statistics PhD students and MS in Statistics students. It is encouraged (but not required) for MS in Applied Statistics and all other students. If you are the latter, you will not get penalized if you get these wrong.

Exercises are from the course textbook *Applied Regression Analysis & Generaized Linear Models, 3rd Edtion (FOX)* --- 14.1, 14.3, 14.6, 15.2, and 15.4.

You can type your answers directly into this document or you can do the work for each exercise on paper, take a picture of your solution, and include the image within this document (work must be legible). 

Alternatively, you can do the Handwork exercises on paper and submit a scanned copy along with your html completing the Data Analysis exercises. That is, you will be submitting a pdf with your solution to the Handwork exercises and an html file with solutions to the Data Analysis exercises. Both documents should be well organized and in the case of the handwork, it must be legible.

### Exercise 14.1

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

:::


### Exercise 14.3

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

:::


### Exercise 14.6

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

:::

### Exercise 15.2

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

:::


### Exercise 15.4

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

:::


## Data analysis

### Exercise D14.1 (Dichotomous)

For this question, we will use the `Chile.txt` dataset, which has a polytomous outcome: voting intention (yes, no, abstain, undecided). For this problem, focus only on the subset of the data with outcomes of either 'yes' or 'no'.

(a) Formulate a model that makes substantive sense in the context of the data set - for example,constructing dummy regressors to represent factors and including interaction regressors where these are appropriate - and fit a linear logistic regression of the response variable on the explanatory variables, reporting the estimated regression coefficients and their asymptotic standard errors.

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::

(b) Construct an analysis-of-deviance table for the model fit in part (a).

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::

(c) Fit a final model to the data that includes the statistically significant effects. Construct an effect display for each high-order term in the model, if your model includes them. If the model is additive, (i) suggest two interpretations of each estimated coefficient; and (ii) construct likelihood-ratio-based 95- percent confidence intervals for the regression coefficients, comparing these with confidence intervals based on the Wald statistic.

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::

(d) Fit a probit model to the data, comparing the results to those obtained with the logit model. Which do you think is better? Why?

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::


### Exercise D14.2 (Polytomous outcome)

Proceed as in Exercise D14.1, but now include all of the data and the four possible outcome values.

Use, as appropriate, one or more of the following: a multinomial logit model; a proportional odds logit model; logit models fit to a set of nested dichotomies; or similar probit models. If you fit the proportional-odds model, test the assumption of parallel regressions. If you fit more than one kind of model, which model do you prefer and why? If you only fit one model, why? Make sure to explain the results and interpretations of the preferred model.

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::


### Exercise D15.3 (GLM Diagnostics)

Return to the logit (and probit) model that you fit in Exercise D14.1.

(a) Use the diagnostic methods for generalized linear models described in this chapter to check the adequacy of the final model that you fit to the data.

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::


(b) If the model contains a discrete quantitative explanatory variable (such as a binned variable), test for nonlinearity by specifying a model that treats this variable as a factor (e.g., using dummy regressors), and comparing that model via a likelihood-ratio test to the model that specifies that the variable has a linear effect. (If there is more than one discrete quantitative explanatory variable, then begin with a model that treats all of them as factors, contrasting this with a sequence of models that specifies a linear effect for each such variable in turn.) Note that this is analogous to the approach for testing for nonlinearity in a linear model with discrete explanatory variables described in Section 12.4.1.

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::


(c) Explore the use of the Cauchy and complementary-log-log links as alternatives to the logit link for this regression. Comparing deviances under the different links, which link appears to best represent the data?

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::


### Exercise D15.1 (Count data)

Long (1990, 1997) investigates factors affecting the research productivity of doctoral students in biochemistry. Long's data (on 915 biochemists) are in the file `Long.txt`. The response variable in this investigation, `art`, is the number of articles published by the student during the last three years of his or her PhD programme. Overview of the explanatory variables are provided in @tbl-long-ex-vars below.

| Variable name   | Definition                                                     |
|:----------------|:---------------------------------------------------------------|
| `fem`           | Gender: dummy variable - 1 if female, 0 if male                |
| `mar`           | Maritial status: dummy variable - 1 if married, 0 if not       |
| `kid5`          | Number of children five years old or younger                   |
| `phd`           | Prestige rating of PhD department                              |
| `ment`          | Number of articles published by mentor during last three years |

: Explanatory variables in `long.txt` data {#tbl-long-ex-vars}

(a) Examine the distribution of the response variable. Based on this distribution, does it appear promising to model these data by linear least-squares regression, perhaps after transforming the response? Explain your answer.

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::


(b) Following Long, perform a Poisson regression of art on the explanatory variables. What do you conclude from the results of this regression? Be sure to interpret the results.

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::


(c) Perform regression diagnostics on the model fit in the previous question. If you identify any problems, try to deal with them. Are the conclusions of the research altered?

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::


(d) Refit Long's model allowing for overdispersion (using a quasi-Poisson or negative-binomial model). Does this make a difference to the results?

::: {.callout-tip icon="false"}
## Solution

YOUR SOLUTION HERE

```{r}


```

:::

